D3064R0: How to Avoid OOTA Without Really Trying

### 2024-04-01 ootavacuous.tex
Add references to table of contents
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Self-review up through the end of Appendix C
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

In order to avoid these rfi links we will substitute rfe
for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe,
where rfe is the set of store-to-load links between different threads.
Any rfi links in a cycle can instead be interpreted as part of sdep.
Although this does shunt additional complexity onto the term
``semantic dependency'', it also enables us to separate cleanly
the interthread and intrathread portions of any given OOTA cycle.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the intuitive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy dependencies, including
semantic ones (depending on one's definition).
That is, even when the potential for a dependency exists in the
source code for a thread, there may be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later in global time than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
We will see examples of this destruction in
Sections~\ref{sec:Global Optimization Can Destroy Dependencies}
and~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies} below.

@@

Consider the following:
\begin{quote}
\begin{verbatim}
x = y - z;
\end{verbatim}
\end{quote}
There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation somehow knows that \co{y} is
always equal to \co{z} at this point then there is no semantic dependency;
the implementation can act as if the statement were simply ``\co{x = 0}''.
We leave aside the question of how the implementation would know this,
given that \co{y} and \co{z} cannot be updated simultaneously
and are subject to change at any time by
other threads (a point we will return to in
Section~\ref{sec:Global Optimization Can Destroy Dependencies}).
% The usual way in the Linux kernel is via a union, but it is probably
% best to ignore that possibility.

@@

The outcome when a given computer executes the machine code in a file
has historically been much better defined than the executions of the
C++ abstract machine.
The hardware's behavior is typically specified with great precision by
the designer or manufacturer, and there are formal, executable memory models
describing exactly what patterns of loads and stores can occur.
Thus, leaving aside questions of asynchronous interrupts and system
calls, the behavior of a CPU executing a particular thread within a
program is entirely determined by the values obtained by the various
memory-load instructions.
(We count read-modify-write instructions as both memory loads and
memory stores.)

@@

A complete C++ implementation consists of much more than just a compiler.
Among other things, for example, it might have a collection of
\co{.h} header files, a linker, runtime libraries, and a dynamic loader.
Nevertheless, for our purposes the compiler is the most important
component because it is what primarily determines the translation from
a C++ source program to directly executable machine code.
We will therefore use the terms ``compiler'' and ``implementation''
somewhat interchangeably.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction or its outputs until the relevant inputs' sources have committed.
The reason is obvious: An input is subject to change at any time until
its source commits to its value,
and the instruction and its outputs can't commit until the CPU has fully
determined what it will do and what they will be.
(On the other hand, the CPU need not wait for inputs that won't affect
the instruction's results.
For instance, a conditional-move instruction wouldn't need to wait for
the source register input to be determined once it knew that the
condition was definitely false.
And, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)

@@

Although it is unexpected, we cannot say that this conclusion is
definitely wrong, because our intuitive notions of semantic
dependency are not clear in cases where multiple loads of the same
variable are present.

@@

For example, nonvolatile relaxed atomic accesses are subject to the
C++ memory model.
The ``order-preserving'' bullet point from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering of quasi-volatile accesses when they are
to the same object,
but such reordering is in any case directly forbidden by the memory model
(the four coherence rules in 6.9.2.2p14--19 \co{[intro.races]}).
The ``onto'' bullet point also largely reiterates restrictions that follow
from the standard, in particular,
that compilers should not invent atomic stores,
duplicate atomic stores, or invent atomic loads.
On the other hand, it is permissible for compilers to omit redundant
atomic stores or fuse nonvolatile atomic accesses of adjacent objects
under appropriate circumstances;
these topics correspond to the ``total'' and ``one-to-one'' bullet points of
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

@@

Acting as if a redundant store was omitted from the source
code cannot create an OOTA cycle, because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of nonvolatile
relaxed atomic stores omitted is a valid execution of the
program with the store present.

@@

This paper assumes that the user defines threads, and that the
C++ implementation executes those threads unchanged.
Future analyses might examine the possibility of ``flattening''
optimizations that combine multiple threads into one.

@@

% @@@ Get citations

Both the ARM and PowerPC architectural memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of any such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering.
% @@@ Cite Grisenthwaite

@@

However, some kernels and applications have ways of handling even
uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one user process, and might react by
killing that user process, taking care to account for memory shared with
other processes.
Some user applications might note that the corrupted data affected only
a particular computation, and might react by restarting that computation.
In both cases, it is only reasonable to react to an uncorrectable ECC
error in a load from a read-only mapped file by re-reading that data
from that file, then restarting that load instruction.
In all cases, the kernel and the user application would need to continue
execution, and would thus require a precise exception.

Systems that offload processing of correctable ECC errors to software
also require precise exceptions.
After all, the software is going to need to be able to correct the error
and then fix up the state to make it appear that the load had returned
the fixed-up value.

>>>>

In order to avoid these rfi links we will substitute rfe
for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe,
where rfe is the set of store-to-load links between different threads.
Any rfi links in a cycle can instead be interpreted as part of sdep.
Although this does shunt additional complexity onto the term
``semantic dependency'', it also enables us to cleanly separate
the interthread and intrathread portions of any given OOTA cycle.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the intuitive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy dependencies, including
semantic ones (depending on one's definition).
That is, even when the potential for a dependency exists in the
source code for a thread, there might be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later in global time than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
We will see examples of this destruction in
Sections~\ref{sec:Global Optimization Can Destroy Dependencies}
and~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies} below.

@@

Consider the following:
\begin{quote}
\begin{verbatim}
x = y - z;
\end{verbatim}
\end{quote}
There appear to be semantic dependencies from \co{y} to \co{x} and from \co{z}
to \co{x}.
However, if the implementation somehow knows that \co{y} is
always equal to \co{z} at this point then there is no semantic dependency;
the implementation can act as if the statement were simply ``\co{x = 0}''.
We leave aside the question of how the implementation would know this,
given that \co{y} and \co{z} cannot be updated simultaneously\footnote{
	At least not by any means within the confines of the standard.}
and are subject to change at any time by
other threads (a point we will return to in
Section~\ref{sec:Global Optimization Can Destroy Dependencies}).
% The usual way in the Linux kernel is via a union, but it is probably
% best to ignore that possibility.

@@

The outcome when a given computer executes the machine code in a file
has historically been much better defined than the executions of the
C++ abstract machine.
The hardware's behavior is typically specified with great precision by
the designer or manufacturer, and there are formal, executable memory models
describing exactly what patterns of loads and stores can occur.
Thus, leaving aside questions of asynchronous interrupts and system
calls, the behavior of a CPU executing a particular thread within a
program is entirely determined by the values obtained by the various
memory-load instructions.
(We count read-modify-write instructions as comprising both memory loads and
memory stores.)

@@

A complete C++ implementation consists of much more than just a compiler.
Among other things, for example, it might have a collection of
\co{.h} header files, a linker, runtime libraries, and a dynamic loader.
Nevertheless, for our purposes the compiler is the most important
component because it is what primarily determines the translation from
a C++ source program to directly executable machine code.
We will therefore use the terms ``compiler'' and ``implementation''
interchangeably.

@@

A CPU may start executing an instruction speculatively, but at some
point it must \emph{commit} to a decision either to definitely execute the
instruction's action with some collection of well defined inputs and outputs
or else to abandon it.
Thanks to the law of cause and~effect, a CPU is not able to commit an
instruction or its outputs until the relevant inputs' sources have committed.
The reason is obvious: An input is subject to change at any time until
its source commits to its value,
and the instruction and its outputs can't commit until the CPU has fully
determined what it will do and what they will be.
(On the other hand, the CPU need not wait for inputs that won't affect
the instruction's results.
For instance, a conditional-move instruction wouldn't need to wait for
the source register input to be determined once it knew that the
condition was definitely false.
And, although no existing architectures do this, there is nothing
in principle to prevent a CPU from committing a \co{multiply}
instruction as soon as either one of its inputs' sources has committed to the
value zero.)
% Are we sure that no CPUs do this?

@@

Although it is unexpected, we cannot say that this conclusion is
definitely wrong, because our intuitive notions of semantic
dependency are not clear in cases where multiple loads of the same
variable are present.
% Can we instead say that transforming the code might require that the
% witness also be transformed?

@@

For example, nonvolatile relaxed atomic accesses are subject to the
C++ memory model.
The ``order-preserving'' bullet point from
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
prohibits reordering of quasi-volatile accesses when they are
to the same object,
but such reordering is in any case directly forbidden by the memory model
(the four coherence rules in 6.9.2.2p14--18 \co{[intro.races]}).
The ``onto'' bullet point also largely reiterates restrictions that follow
from the standard, in particular,
that compilers should not invent atomic stores,
duplicate atomic stores, or invent atomic loads.
On the other hand, it is permissible for compilers to omit redundant
atomic stores or fuse nonvolatile atomic accesses of adjacent objects
under appropriate circumstances;
these topics correspond to the ``total'' and ``one-to-one'' bullet points of
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

@@

Acting as if a redundant store was omitted from the source
code cannot create an OOTA cycle, because any valid execution (OOTA or not)
of the program with the first of a back-to-back pair of nonvolatile
relaxed atomic stores omitted is a valid execution of the
program with the store present.
% This is awkward because a reader might reasonably extend the "Acting
% as if" all the way to the comma.  The best I could do was "It is not
% possible to create an OOTA cycle by acting as if a redundant store
% was omitted from the source code", which is admittedly not all that
% great.

@@

This paper assumes that the user defines threads, and that the
C++ implementation executes those threads unchanged.
Future analyses might examine the possibility of ``flattening''
optimizations that combine multiple threads into one, but note
that such optimizations are more difficult than they first might
appear.

@@

Both the ARM and PowerPC architectural memory models permit reordering prior loads
against subsequent stores, but actual tests on recent hardware fail to
produce any evidence of such reordering.
Nevertheless, thus far neither ARM nor PowerPC hardware architects
have been willing to strengthen their memory models so as to forbid
such reordering, despite a number requests to do
so~\cite{RichardGrisenthwaite2023TightenRelaxedJustSayNo}.

@@

However, some kernels and applications have ways of handling even
uncorrectable errors.
Operating-system kernels encountering an ECC error might note that the
affected data was being used by only one user process, and might react by
killing that user process, taking care to account for memory shared with
other processes.
Some user applications might note that the corrupted data affected only
a particular computation, and might react by restarting that computation.
On the other hand, it is only reasonable to react to an uncorrectable ECC
error in a load from a read-only mapped file by re-reading that data
from that file, then restarting that load instruction.
In such cases, the kernel and the user application would need to continue
execution, and would thus require a precise exception.

Systems that offload processing of correctable ECC errors to software
also require precise exceptions.
After all, the software is going to need to be able to correct the error
and then fix up the state to make it appear that the load had returned
the fixed-up value.
% I was tempted to add a sentence saying that although one could imaging
% ECC on memory but not on caches, today's caches are a bit large for this
% to be a winning strategy.  One might nevertheless consider omitting ECC on
% the smallest caches closest to the CPU, but this is unlikely to allow LB.
